Charack: tool for real-time generation of pseudo-infinite virtual worlds for 3D games

### 2009-07-26 paper/paper.tex
- Advisor's release.

\begin{abstract}
In MMO games the player's experience is mainly influenced by the size and details of the virtual world. Technically the bigger the world is, the bigger is the time the player takes to explore all the places. This work presents a tool able to generate pseudo-infinite virtual worlds with different types of terrain. Using a combination of algorithms and content management methods, the tool is able to create beaches, islands, bays and cost lines that imitates the real world landscapes. One of the contributions of the tool is the ability to generate giant pieces of land focusing on the coast line generation. The development of the present work aimed to handle separately the generation of content for all elements in the world (continents, terrains, etc.). The main point in the work is the coast line generation, not the content inside the continents.
\end{abstract}

@@

The computer games market has been evolving considerably over the years. Since the first console, the hardware performance has increased and new graphic technologies were developed, resulting in a wide range of themes and game styles. In the multiplayer games, players interact with other human beings and also with NPCs, which are represented by virtual characters. The kind of game is popular and the social interaction between players is a matter of research ~\cite{onlineGaming,ducheneaut}. In the category of multiplayer games there are the massively multiplayer on-line (MMO) ones, which are on-line games featuring large number of players interacting with each other in a persistent virtual world.

An MMO can feature millions of players, such as EverQuest ~\cite{EverQuest} and World of Warcraft ~\cite{Blizzard}, the latter with more than 6 million subscribers ~ \cite{vicioEstrutura}. A persistent virtual world is an important topic to keep the game fun and attractive to the player. The bigger is the world to be explored, technically the bigger is the time the player has to spend in order to explore all the places. As a result of such huge virtual worlds, the creation (and subsequent upgrade) of those worlds is a complex task. EverQuest and World of Warcraft present a virtual world with a wide diversity of geographical features such as mountains, valleys, forests, fields, caves, etc.., and the vast majority of them nominated and related the history of the game. The manual creation of those virtual worlds requires a team able to create heightmaps, adorn landscapes, ensure usability of the map (avoid unreachable places, for example), create interesting place for players, etc. To help on that task, the development of a tool able to generate complex virtual worlds is useful to speed up the development of 3D games such as MMOs.

The solution proposed in this work is a tool able to generate complex virtual worlds in real-time using noise-based techniques of terrain generation. The proposed tool is able to generate complete virtual world featuring continents, oceans, flat areas, mountains, etc, with all those elements generated in real time, on demand as the player moves along the world.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The tool was designed to allow developers to use its features in order to generate 3D terrains for games, particularly MMOs, with minimal human intervention in the generation process. The size of the resulting world is customizable, but it can reach billions of pixels. The use of the term "pseudo" is necessary due physical limitations in computers hardware: an unsigned integer, for instance, can store a certain amount of data; if there were no physical limitations, the tool would be able to generate, in fact, an infinite world.

The content generation is made on demand. As the user moves along the world, the elements inside the user's view are processed and stored into the memory and the ones away from the user's view are removed. Even though the generation of all elements is based on random calculations, if the player visits an specific point {\tt A}, then walks for miles generating a completely different set of landscapes, and returns to point {\tt A}, the same previously seen landscape will be shown again.

Besides the heightmap generation, the tool is also able to create oceans and continents, all of them customizable. Unlike other related works which deal with the generation of continents as a consequence of the heightmap generation, this paper presents as its main contribution a new approach for content generation in virtual worlds. In this approach, the generation of continents, topography and coastlines are handled separately.

The shape of continents is defined on a macro level as a result of an algorithm for maps generation that work projecting pixels onto a sphere. The coast line generation is performed on a micro level. As a result, the tool is able to produce cliffs, beaches, bays and islands.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related work}
\label{sec:related-work}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%As seções seguintes apresentam diversas abordagens utilizadas para a criação de mundos virtuais finitos e pseudo-infinitos, bem como formas de geração de conteúdo para eles (criação de relevo).

\subsection{Virtual city}

The procedural content generation is an old topic in the field of computer graphics. The application of such technique to generate a complete virtual world was used by ~\cite{infinicity}, whose goal was to create a virtual city that was visually interesting and complex.

In the approach the world were divided into a grid consisting of several squares, called cells. The location of each cell is used with a global seed as an input for a hash function ~\cite{wangHash}. The result of this function is used as a seed for a pseudo random number generator and it will define all the characteristics of the buildings within a cell. As a consequence of that approach the contents of a cell is always the same, no matter how far the user moves or if the cell is removed from memory.

To ensure acceptable use of memory and CPU, only the elements inside the user's view are generated. As the user moves through the city, new cells are added to the user's view and the new content is generated. When the cell leaves the user's view, it is removed from memory and its resources are released. The cells are placed in square loops around the user and are considered inside the user's view if it is close enough and if it is located within a 120° view range.

\subsection{Layered generation and texturization}

Another related work was a tool used to build the SkyCastle multiplayer game engine~\cite{LandscapeGeneration,SkyCastle}. For the heightmap generation, parameterized procedures and fractal based systems are combined in layered approach: starting with a base map, the application merges a new map with to the base in each iteration. The new maps are pre-calculated and generated using Perlin noise.

One of the texturization methods presented are the use of a large image able to cover the whole world. Although this approach is useful for small scenes, it is not suitable for large scenes nor virtual worlds, since the image size could reach prohibitive proportions. To avoid this problem, an reticulate approach is suggested ~\cite{CSHD03}; using that approach a texture cell is created so that the a combination of them produces a smooth and continuous and textured plane. The results are acceptable, but it is not visually attractive to the user, because the landscape features an unusual pattern that will never exist in the real world. In order to achieve a better visual result, an approach proposed by ~\cite{LN03} is suggested. In that approach, a set of pre-defined border textures is combined with reticulate textures. The idea is to apply the border textures on an already textured plane, but using transparency in the border textures. As a consequence, border textures will overlap the some regions of the plane and produce a less homogeneous landscape, which creates a better visual result.

In order to decorate the virtual world, three procedural plant generation methods are used: L-System ~\cite{PL90}, component-based generation ~\cite{LD98} and parameterized trees ~\cite{WP95}.

\subsection{Procedural planets}

The use of fractals and their combinations with other methods are also used to procedurally generate planets ~\cite{LindaOndrej2007}. In this approach, the generated world is not infinite, but it is spherical and simulates the view of the planet Earth. Starting with a recursive subdivision of an octahedron, a spherical world is created and it is used as a first step to several algorithms of heightmap generation. The heightmap is obtained as a result of several techniques, each one featuring its peculiarities and results. Among the techniques used are: generation using random failures, midpoint displacement (including a multifractal variation) and Perlin noise (and its many variations).

\subsection{Erosion fractals}

Another approach for procedural terrain generation in real time uses fractals affected by erosion ~\cite{terrain_generation}. As pointed out, the increasing processing power of home computers has enabled games to make use erosion simulations almost in real time. As a consequence, an approach involving fractals and erosion is possible to be reached, resulting in more realistic terrain. In order to achieve that, a heightmap was created from a Voronoi diagram, where each vertex of the diagram is called {\it functionality point}; the value of each cell in the heightmap is obtained through a linear combination of distances of the closest functionality points. Once the heightmap is generated, it is combined with a noise algorithm in order to produce mountains. The sharp lines introduced by the well-defined Voronoi diagram are removed with the application of a turbulence filter. For the erosion simulation an hybrid method is used, which is a combination of two techniques: the thermal ~\cite{MusgraveErosao} and the hydraulic erosion.

\subsection{Stochastic subdivision of a quadtree}

Another related work uses stochastic subdivision techniques in order to generate a virtual world ~\cite{dollins-thesis}. The main goal was to create a virtual world of large proportions, however the creation of its content is procedurally made and performed on-demand with a multi-resolution approach. For the heightmap generation a recursive sub-division process is applied to a quadtree. The height of each vertex is defined by the midpoint of each cell in the quadtree. For each level of subdivision, the midpoint of the each new child cell is defined by the parent cells around the cell being divided. The height of all other vertexes, such as the one at the corners, is defined as a result of an interpolation of the closest midpoints. In each level of detail, more cells are divided and replaced with the new generated child cells.

\subsection{Map generator}
\label{sec:geradortorben}

Another related work focusing on the generation of maps for planets, which is based on the generation of spheric worlds using a recursive subdivision of a tetrahedron ~\cite{torben}. Although the work is not presented as a graphical tool where the user can explore a 3D environment, all generated information is part of a complete virtual world featuring highly customizable continents and oceans. The presented idea is to create a map as a result of a projection of pixels onto a sphere, a method similar to ray tracing~\cite{raytracing}. In the same way it is performed in ray tracing, where rays leaving the camera collide with something and create a visible pixel, each pixel of the map being generated is projected onto sphere. The projection algorithm sets a height value to each pixel found, resulting in a complete heightmap.
In order to render the map, initially the visible points on the sphere surface are found. After that, the height value for each one of those points is calculated as follows:

\begin{itemize} 
  \item Insert the sphere inside the tetrahedron.
  \item Divide tetrahedron into two smaller tetrahedral. 
  \item Choose which of the two tetrahedral contains the point being analyzed. 
  \item Repeat the above steps until the tetrahedron is small enough.
  \item Use the average height of the each tetrahedron vertex in order to calculate the height of the point being analyzed. 
\end{itemize}

Initially the sphere is placed inside a irregular tetrahedron and information about height and a random seed are set for each of the vertexes. After the tetrahedron is divided in two pieces by a plane formed between the midpoint of the longest edge and the two end points of the edge in front of the midpoint of the longest edge. In short, each of the two resulting tetrahedral have three points from the original tetrahedron and a new vertex placed in the midpoint of the largest edge. The height and the seed of the new vertex is calculated from the height and seed of the end points of longest edge. The process is repeated recursively for each of the tetrahedral until it is small enough to contain only the desired point.

@@

The following sections explain in a top-down approach how the tool was structured to achieve the described goals.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Initial analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

There are several related works concerning the generation of virtual worlds, whether finite or infinite. One of the related work creates an infinite city ~\cite{infinicity} that is presented to the user on demand as it walks on the ground. This work was the ground zero for Charack development, however the original idea was changed in order to make the tool suitable to generate more types of terrains (mountains, plains, continents, etc.), not only streets and buildings. The approach of content generation made on demand was maintained.

The procedurally generated world approach ~\cite{LindaOndrej2007} is very close to the concept of content generation aimed for Charack. In that work, a spherical planet is created as a result of a recursive division of a geometric shape, then noise functions are applied to the mesh to generate the heightmap. There is no distinction between the content generation approach for continents and the content generation for the lands within the continents. As a result the continents are created by flooding the heightmap with a water plane, which will produce the coast lines based on height of the sea level and the amount of ripples in the topography. The content itself is not generated on demand. Charack was created from an evolution of this idea, but subject to some limitations. The world created by Charack handles differently the generation of continents, coast lines and the heightmap within them. Each of these generation processes has its own peculiarities, which can be tweaked in order to customize specific details in the world with no interference in the other processes. This is the case of the creation of long beach on the coast without changing the basic structure or the topography of the continent that the beach is inserted. Charack also generates the content on demand, however it does not use a spherical approach, the resulting world contained within a plan.

The generation of a virtual world as a result of recursive subdivisions of a quadtree ~\cite{dollins-thesis} is very similar to the Charack proposal. In that work, a world with huge proportions is created and its content is generated on demand as the user moves. The heightmap is created in a parameterized and multi-resolution way, so the closer the user of place, the greater is the amount of detail there. There is also no distinction in the generation process of continents/coastlines/land content. The heightmap generation proposed is used by Charack, however continents and the coastlines generation process are completely different.

Analyzing the related work, virtual worlds are generated several approaches, but in none of them handles differently the content generation for continents, coastlines and topography. Although there are variations in how the heightmap is created, the generation of continents is a result of a water flooding plane. This approach allows the developers to focus on the content generation for the land, however it has a simple approach concerning continents and coastlines. The main idea and contribution of Charack is the content generation handled differently for each world element (continent, coastline, etc.), with an aggressive and specific approaches for each one. Most of the related work focuses on heightmap generation, so we believe the Charack contribution will be better if the efforts are focused on the generation of continents and coastlines.

@@

In order to create a virtual world that reaches the presented proposal, a top-down approach is used for the content generation. The Charack data flow begins in a macro view of the world, which are the continents, evolving to a micro view of the planet, which are the generation of content for each vertex that will be drawn in the screen. Figure ~\ref{fig:planejamento_estrutura_basica} shows Charack basic structure.

@@

At the top of the chain is the maps generator, which creates the continents that exist throughout the virtual world. This module is an encapsulation of the solution created by ~\cite{torben}, which was described in the section ~\ref{sec:geradortorben}. When the tool is initiated, it uses a user defined seed to generate all the continents. Once the continents are generated, all the information related to terrain types (land, water and coast) are stored in a matrix, called {\bf macro-matrix (MM)}, which is used by all the other algorithms.

@@

Below the MM and the maps generator is the slice manager (SM). It extracts a portion of the virtual world (the user's view) and provide the render engine with information about the heightmap. In order to obtain the required information to create the heightmap, the slice manager uses the coastline generator (CG), which uses height generator (HG) and the data stored in the MM.

In the context of the SM, there is no information about land or water, it only knows a set of pixels of the virtual world and its height values. Using the position of the user as a guide, the SM slices virtual world and, for each collected pixels, it queries the CG in order to find out its height value.

@@

The coastline generator (CG) will map each pixel of the slice manager to the MM in order to find out the terrain type of that pixel. If the pixel being analyzed is mapped to a location in the MM that is described as water, then the CG assigns a height value equals to sea level for the pixel and returns it to the SM. If the pixel is mapped to a place described as (simple) land, then the CG will use the information provided by the HG in order to find out the height value for that pixel. Finally, if the pixels is mapped to a place described as coast, then the CG uses its own structure (together with the MM) to set height value for the that pixel.

The resulting virtual world technically has height and width defined by the maximum size of a signed integer. It is physically impossible to generate a MM with such proportions. Since the MM is smaller than the virtual world, an MM's entry {\tt (i, j)} represents several pixels in the virtual world. Figure \ref{fig:mapeamento_macro} illustrates the MM mapping process.

@@

The smaller is the MM size, the more pixels in the virtual world will be represented by the same entry in the MM. If the virtual world had {\tt 1000x1000} as its size and the MM had {\tt 10x10} as its size, for instance, it means that for each pixel of the MM represents 100 pixels in the virtual world. If one of these pixels is described in the MM as coast, then there is an area of {\tt 100x100} pixels in the virtual world that must be a coast. To solve this problem the CG acts. When any of those pixels in that particular area of the virtual world is analyzed by the CG, it will work on it and return it with different values, which will result in a coast line for that area, not am area entirely filled with land or water.

@@

The rendering engine draws created heightmap in the screen. The result is rendered as triangles mesh that is textured according to the height value of each vertex in the mesh.

@@

The main problem concerning Charack's implementation was the on demand content generation. Based on the fact the user can only see what is inside the view area, all the content generation algorithms need to take into account {\it only} the information that is available in the user's view. Even though this approach is efficient for resources management (process only the visible elements), it increases the complexity of the content generation algorithms.

The algorithm that generates mountains, for instance, has no way to determine where the mountain ends, because the world outside the user's view technically does not exist yet, it will be generated as the user moves. One approach to solve that problem would be the use of a function that describes the mountain chain, but this function should not rely on begin/end points, because they could not exist in a certain time. If that function does not need any begin/end points, at least it would have to rely on the position of the user in the virtual world. If the function must be aware of some special points, those points have to be previously processed, which would break the on demand content generation concept.

@@

The solution makes the player able to look at the screen and see a slice of the virtual world, but with no explicit divisions in the world, such as cells. Unlike what was done in ~\cite{infinicity} where the user's view is a cone, the Charack user's view is a square centered on the user. Using the user position {\tt (x, y, z)}, the visible content around that coordinate is sliced from the world and drawn in screen. If the player reaches a place or a complete slice is not possible, such as the world boundaries, no content is displayed beyond the world area.

In order to texture the sliced data a set of images are interpolated and managed by the shading language GLSL ~\cite{glsl}. The height value of the pixel defines the interpolation weight of each texture. As a consequence, a sand texture has higher weight for pixels featuring a low height value, for instance. Figure ~\ref{fig:set_texturas} shows all the available textures.

@@

The generation of continents and oceans has been proposed in order to break the monotony of a landscape composed only by land and to increase the similarity of the virtual world with the real world landscapes. The solution for the continents generation consist in pre-process the land areas and store that information for other tool calculations. With that approach, the on demand content generation has been partially broken, since the continents are generated before all the other content, but it ensures a better control over water/land areas.

The continents generation is based on the approach described in section ~\ref{sec:geradortorben}. That planet generator was used because it has several parameterization options, such as the possibility to use a seed to manage all the random calculations. Figure ~\ref{fig:torben} {\it (a)}, {\it (b)} and {\it (c)} illustrates the results obtained with the planet generator of  ~\cite{torben}.

@@

As previously explained in section ~\ref{sec:basic-structure}, each pixel of the MM is mapped to several pixels in the virtual world. A direct consequence of that mapping process is the generation of large areas featuring straight land lines. If there were no hardware limitation and if it were possible the generation a matrix with the exact size of the virtual world, the matrix would contain the necessary resolution for the tool to accurately determine whether a pixel is land or not land, in a ratio of 1:1 (one MM pixel is mapped to one world pixel). This approach, however, is not suitable because a matrix with such proportions consumes many resources and processing time. Although the tool allows customization of the MM size, tests presented that a {\tt 800x800} pixels MM has is quite enough information to be processed by all the other algorithms of the tool.

Figure \ref{fig:ilustracao_mapeamento_macro} illustrates the results obtained by the tool when no algorithm is used to generate extra content to fill the rectilinear spaces of the virtual world.

@@

This figure illustrates a place in the virtual world that represents the transition between two different points of the MM (a land point and a water point). To explain what is happening, assume the tool is drawing the world at position {\tt (x, y, z)}, which is the mapping result of a pixel {\tt (i, j)} in the MM, which is described as land; as the tool increases the coordinate in order to draw the landscape, each new position is mapped to the MM. If the result of the mapping process of new coordinate, {\tt (x + 1, y, z))} for instance, is still the point {\tt (i, j)} in the MM, then the tool will again draw a land pixel on the screen. Assuming that only at point {\tt (x + 10, y, z)} the pixels start being mapped to a different pixel in the MM, such as {\tt (i + 1, j)} (and {\tt (i + 1, j)} is described as land), then all points before the position {\tt (x + 10, y, z)} are drawn as water and all pixels after that location are drawn as land.

The figure shows clearly when the world coordinates start being mapped to a different entry in the MM, which is when the tool replaces the land rendering with water rendering. As a consequence of no algorithm being applied to generate content for that transition area, the user will move along the coastline and will see only straight lines.

@@

The mapping process of the pixels of the virtual world to the MM produces very unrealistic landscapes. A real world beach has a natural curvature and hardly have a length of 20 Km in a perfectly straight configuration, as the beaches generated by Charack. Although the objective of this work is not to create photo-realistic landscapes, such unreal beaches are not acceptable. To circumvent this problem, a coastline disturbance is applied to the locations where the mapping process is made between two MM points, one of them described as land and the other one described as water. The algorithm is described below.

The MM has a full description of what is land and what is water in the virtual world. Each of its pixels has a descriptor, which tells the other algorithms what type of terrain one pixel of the virtual world is after it is mapped to the MM. Charack features three types of terrain: water, land (continent) and offshore (land in contact with water). After the continents are pre-processed and stored in the MM, it only features information about land (continents) and water.

From that moment, the first step of the coastline disturbance algorithm is performed. Using the current MM as its input, the algorithm scan each MM's pixel, updating the descriptor of pixels that represent a coast. A pixel is said to be coast when at least one of its neighbors is water. After the algorithm end, the MM contains the three types of terrain described before (water, land and offshore). The next step to apply disturbance to the coastline is the content generation based on the descriptor of each pixel in the MM. When the tool is creating content to draw on the screen, each pixel being drawn is tested against its descriptor in the MM. If the pixel is mapped to a land pixel in the MM, then the function will set a height value for that point. If the pixel is mapped to a water pixel in the MM, then the function will set the sea level height to that point. Finally if the pixel is mapped to a offshore pixel in the MM, then the function will disturb the land/water information of that mapping process, which will result on a non-straight coast line. Figure ~\ref{fig:funcionamento_quebra_linearidade} illustrates the algorithm.

>>>>

\begin{abstract}
In MMO games the player's experience is mainly influenced by the size and details of the virtual world. Technically the bigger the world is, the bigger is the time the player takes to explore all the places. This work presents a tool (named Charack) able to generate pseudo-infinite virtual worlds with different types of terrains. Using a combination of algorithms and content management methods, Charack is able to create beaches, islands, bays and coastlines that imitates real world landscapes. The tool clearly distinguish the generation of the continents, terrains, coastline and vegetation. The contributions of the tool is the ability to generate giant pieces of land (or landscape) focusing on detailed coastline generation, by means of using procedural algorithms.
\end{abstract}

@@

The computer games market has been evolving considerably over the years. Since the first console, the hardware performance has increased and new graphic technologies were developed, resulting in a wide range of themes and game styles. In the multiplayer games, players interact with other human beings and also with NPCs, which are represented by virtual characters. That kind of game is popular and the social interaction between players is a matter of research ~\cite{onlineGaming,ducheneaut}. In the category of multiplayer games there are the massively multiplayer on-line (MMO) ones, which are on-line games featuring large number of players interacting with each other in a huge virtual world.

An MMO can feature millions of players, such as EverQuest ~\cite{EverQuest} and World of Warcraft ~\cite{Blizzard}, the latter with more than 6 million subscribers ~ \cite{vicioEstrutura}. A persistent virtual world is an important topic to keep the game fun and attractive to the player. The bigger is the world to be explored, technically the bigger is the time the player has to spend in order to explore all the places. As a result of such huge virtual worlds, their creation and subsequent upgrade are a complex task. EverQuest and World of Warcraft present a virtual world with a wide diversity of geographical features such as mountains, valleys, forests, fields, caves, etc, and most of them have specific names  and are related to the game history. The manual creation of those virtual worlds requires a team able to design heightmaps, adorn landscapes, ensure usability of the map (avoid unreachable places, for instance), create interesting places for players, etc. To help on that task, the development of a tool able to generate complex virtual worlds is useful to speed up the development of 3D games such as MMOs.

The solution proposed in this work is the development of a tool, called Charack, able to generate complex virtual worlds in real-time using noise-based techniques for terrain generation. Charack was designed to allow developers to use its features in order to generate 3D terrains for games, particularly MMOs, with minimal human intervention in the generation process. The content generation is made on demand. As the user moves along the world, the elements inside the user's view are processed and stored into the memory and the ones away from the user's view are removed. Even though the generation of all elements is based on random calculations, if the player visits an specific point {\tt A}, then walks for miles generating a completely different set of landscapes, and returns to point {\tt A}, the same previously seen landscape will be shown again.

Charack handles separately the content generation of continents, topography and coastlines, so each of those elements can be independently adjusted in order to produce highly customized results. The use of the term "pseudo" is necessary due physical limitations in computers hardware: an unsigned integer, for instance, can store a certain amount of data; if there were no physical limitations, the tool would be able to generate, in fact, an infinite world.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related works}
\label{sec:related-work}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

There are several related works concerning the generation of finite or infinite virtual worlds. One of them creates an infinite city that is presented to the user on demand as it walks on the ground~\cite{infinicity}. The world were divided into a grid composed of several squares, called cells. The location of each cell is used with a global seed as an input for a hash function ~\cite{wangHash}. The result of this function is used as a seed for a pseudo random number generator and it defines all the characteristics of the buildings within a cell. As a consequence of that approach the contents of a cell is always the same, no matter if the user moves and that cell is removed from the memory. That work was the ground zero for Charack development, however the original idea was changed in order to make the tool suitable to generate more types of terrains (mountains, plains, continents, etc.), not only streets and buildings. The approach of content generation made on demand was maintained, but the cells organization was removed.

The use of a procedurally generated world approach ~\cite{LindaOndrej2007} is very close to the concept of content generation aimed for Charack. In that work, a spherical planet is created as a result of a recursive division of a geometric shape, then noise functions are applied to the mesh to generate the heightmap. There is no distinction between the content generation approach for continents and the content generation for the terrain within the continents. As a result the continents are created by flooding the heightmap with a water plane, which will produce the coastlines based on the sea level height and the amount of ripples in the topography. The content itself is not generated on demand. Charack was created from an evolution of that idea, but with limitations. The world created by Charack handles differently the content generation of continents, coastlines and the heightmap and it also generates the content on demand, however it does not use a spherical approach.

Another related work was a tool used to build the SkyCastle multiplayer game engine~\cite{LandscapeGeneration,SkyCastle}. For the heightmap generation, parameterized procedures and fractal based systems are combined in a layered approach: starting with a base map, the application merges a new map with the base one in each iteration. The new maps are pre-calculated and generated using Perlin noise. To texturize the landscape and to adorn it with plants, several techniques are used ~\cite{CSHD03,PL90,LD98,WP95}. Charack uses a similar noise approach in order to create the terrain height, however it was not initially designed to generate extra content such as trees and plants.

The generation of a virtual world as a result of recursive subdivisions of a quadtree ~\cite{dollins-thesis} is very similar to the Charack proposal. In that work, a world with huge proportions is created and its content is generated on demand as the user moves. The heightmap is created in a parameterized and multi-resolution way, so the closer the user is of place, the greater is the amount of detail there. There is also no distinction in the content generation process of continents, coastlines and land. The proposed heightmap generation is used by Charack, however continents and the coastlines generation process are completely different.

Another approach uses fractals affected by erosion for real time, procedural generation of terrains ~\cite{terrain_generation}. For the erosion simulation, thermal ~\cite{MusgraveErosao} and hydraulic methods are used. Charack has no feature connected with the weather influence, even though it produces some sort of very basic erosion simulation when all sharp edges of the heightmap are removed by a smoothing algorithm.

The planet map creator based on the generation of a spheric world using a recursive subdivision of a tetrahedron is another approach concerning procedural content generation~\cite{torben}. All the generated information is part of a complete virtual world featuring highly customizable continents and oceans created as a result of a projection of pixels onto a sphere, a method similar to ray tracing~\cite{raytracing}. That planet map creator is used by Charack as a starting point on the continent generation process.

@@

@@

Analyzing the related works, virtual worlds are generated through several approaches, but none of them handles differently the content generation for continents, coastlines and topography. Although there are variations in how the heightmap is created, the generation of continents is a result of a water flooding plane. This method allows the developers to focus on the content generation for the land, however it has a simple approach concerning continents and coastlines. The main idea and contribution of Charack is the content generation handled differently for each world element (continent, coastline, etc.), with an aggressive and specific approach for each one. Most of the related work focuses on heightmap generation, so we believe the Charack contribution will be better if the efforts are focused on the generation of continents and coastlines.

In order to create a virtual world that reaches the presented proposal, a top-down plan is used for the content generation. The Charack data flow begins in a macro view of the world, which are the continents, evolving to a micro view of the planet, which are the content generation for each vertex that will be drawn in the screen. Figure ~\ref{fig:planejamento_estrutura_basica} shows Charack basic structure.

@@

At the top of the chain is the map generator, which creates the continents that exist throughout the virtual world. This module is an encapsulation of the solution created by ~\cite{torben}. When the tool is initiated, it uses a user defined seed to generate all the continents. Once the continents are generated, all the information related to terrain types (land, water and coast) are stored in a matrix called {\bf macro-matrix (MM)}, which is used by all the other algorithms.

@@

Below the MM and the map generator is the slice manager (SM). It extracts a portion of the virtual world (the user's view) and provide the rendering engine with information about the heightmap. In order to obtain the required information to create the heightmap, the slice manager uses the coastline generator (CG), which uses the height generator (HG) and the data stored in the MM.

In the context of the SM, there is no information about land or water, only a set of vertexes and their height values. Using the position of the user as a guide, the SM slices the virtual world and, for each collected pixel, it queries the CG in order to find out the height value for that pixel.

@@

The coastline generator (CG) will map each pixel of the slice manager to the MM in order to find out the terrain type of that pixel. If the pixel being analyzed is mapped to a location in the MM that is described as water, then the CG assigns a height value equals to sea level for the pixel and returns it to the SM. If the pixel is mapped to a place described as (simple) land, then the CG will use the information provided by the HG in order to set the height value for that pixel. Finally, if the pixels is mapped to a place described as coast, then the CG uses its own structure (together with the MM) to set the height value for the that pixel.

The resulting virtual world technically has height and width defined by the maximum size of a signed integer, however it is physically impossible to generate a MM with such proportions. Since the MM is smaller than the virtual world, an MM's entry {\tt (i, j)} represents several pixels in the virtual world. Figure \ref{fig:mapeamento_macro} illustrates the MM mapping process.

@@

The smaller is the MM size, the more pixels in the virtual world will be represented by the same entry in the MM. If the virtual world had {\tt 1000x1000} as its size and the MM had {\tt 10x10} as its size, for instance, it means that each entry of the MM represents 100 pixels in the virtual world. If one of these entries is described in the MM as coast, then there is an area of {\tt 100x100} pixels in the virtual world that must be a coast. When any of those pixels in that particular area of the virtual world is analyzed by the CG, it will work on it and return it with different values, which will result in a coastline for that area, not an area entirely filled with land or water.

@@

The rendering engine draws the created heightmap in the screen. The result is rendered as a triangles mesh that is texturized according to the height value of each vertex in the mesh.

@@

The main problem concerning Charack's implementation was the content generation performed on demand. Based on the fact the user can only see what is inside the visible area, all the content generation algorithms need to take into account {\it only} the information that is available in the user's view. Even though this approach is efficient for resources management (process only the visible elements), it increases the complexity of the content generation algorithms.

The algorithm that generates mountains, for instance, has no way to determine where the mountain ends, because the world outside the user's view technically does not exist yet, it will be generated as the user moves. One approach to solve that problem would be the use of a function that describes the mountain backbone, but this function should not rely on begin/end points, because they could not exist in a certain time. If that function does not need any begin/end points, at least it would have to rely on the position of the user in the virtual world. If the function must be aware of some special points, those points have to be previously processed, which would break the on demand content generation concept.

@@

The solution makes the player able to look at the screen and see a slice of the virtual world, but with no explicit divisions in the world, such as cells. Unlike what was done in ~\cite{infinicity} where the user's view is a cone, the Charack user's view is a square centered on the user. Using the user position {\tt (x, y, z)}, the visible content around that coordinate is sliced from the world and drawn in the screen. If the player reaches a place where a complete slice is not possible, no content is displayed beyond the world area.

In order to texturize the sliced data a set of images are interpolated and managed by the shading language GLSL ~\cite{glsl}. The height value of the pixel defines the interpolation weight of each texture. As a consequence, a sand texture has higher weight for pixels featuring a low height value, for instance. Figure ~\ref{fig:set_texturas} shows all the available textures.

@@

The generation of continents and oceans has been proposed in order to break the monotony of a landscape composed only by land and to increase the similarity of the virtual world with real world landscapes. The solution for the continents generation consists in pre-process the land areas and store that information for later calculations. With that approach, the on demand content generation has been partially broken, since the continents are generated before all the other content, but it ensures a better control over water/land areas.

The continents generation is based on the planet map creator described in section ~\ref{sec:related-work}. That planet generator was used because it has several parameterization options, such as the possibility to use a seed to manage all the random calculations. Figure ~\ref{fig:torben} {\it (a)}, {\it (b)} and {\it (c)} illustrates the results obtained with the planet generator of  ~\cite{torben}.

@@

As previously explained in section ~\ref{sec:basic-structure}, each entry of the MM is mapped to several pixels in the virtual world. A direct consequence of that mapping process is the generation of large areas featuring straight land lines. If there were no hardware limitation and if it were possible to generate a MM featuring the exact size of the virtual world, the matrix would contain the necessary resolution for the tool to accurately determine whether a pixel is land or not land, in a ratio of 1:1 (one MM entry is mapped to one world pixel). This approach, however, is not suitable because a matrix with such proportions consumes many resources and processing time. Although the tool allows customization of the MM size, tests presented that a MM featuring {\tt 800x800} as its size has enough information to be processed by all the other algorithms.

@@

Figure \ref{fig:ilustracao_mapeamento_macro} illustrates the results obtained by the tool when no algorithm is used to generate extra content to fill the rectilinear spaces of the virtual world. This figure illustrates a place in the virtual world that represents the transition between two different points of the MM (a land point and a water point). To explain what is happening, assume the tool is drawing the world at position {\tt (x, y, z)}, which is the mapping result of an entry {\tt (i, j)} in the MM, which is described as land; as the tool increases the coordinate in order to draw the landscape, each new position is mapped to the MM. If the result of the mapping process of teh new coordinate, {\tt (x + 1, y, z))} for instance, is still the entry {\tt (i, j)} in the MM, then the tool will again draw a land pixel on the screen. Assuming that only at point {\tt (x + 10, y, z)} the pixels start being mapped to a different entry in the MM, such as {\tt (i + 1, j)} (and {\tt (i + 1, j)} is described as land), then all points before the position {\tt (x + 10, y, z)} are drawn as water and all pixels after that location are drawn as land.

Figure \ref{fig:ilustracao_mapeamento_macro} shows clearly when the world coordinates start being mapped to a different entry in the MM, which is when the tool replaces the land rendering with water rendering. As a consequence of no algorithm being applied to generate content for that transition area, the user will move along the coastline and will see only straight lines.

@@

The mapping process of the pixels of the virtual world to the MM produces very unrealistic landscapes. A real world beach has a natural curvature and is not likely to have a length of 20 Km in a perfectly straight configuration, as the beaches generated by Charack. Although the objective of this work is not to create photo-realistic landscapes, such unreal beaches are not acceptable. To circumvent this problem, a coastline disturbance is applied to the locations where the mapping process is made between two MM points, one of them described as land and the other one described as water. The algorithm is described below.

The MM has a complete description of what is land and what is water in the virtual world. Each of its entries has a descriptor, which tells the other algorithms what type of terrain one pixel of the virtual world is after it is mapped to the MM. Charack features three types of terrain: water, land (continent) and offshore (land in contact with water). After the continents are pre-processed and stored in the MM, it only contains information about land (continents) and water.

From that moment, the first step of the coastline disturbance algorithm is performed. Using the current MM as its input, the algorithm scan each MM's entries updating the descriptor of all entries of representing coasts. A entry is classified as coast when at least one of its neighbors is water. After the algorithm ends, the MM contains the three types of terrain described before (water, land and offshore). The next step to apply disturbance to the coastline is the content generation based on the descriptor of each MM entry. When the tool is creating content to draw on the screen, each pixel being drawn is tested against its descriptor in the MM. If the pixel is mapped to a land entry in the MM, then the function will set a height value for that point. If the pixel is mapped to a water entry in the MM, then the function will set the sea level height to that point. Finally if the pixel is mapped to a offshore entry in the MM, then the function will disturb the land/water information of that pixel, which will result on a non-straight coastline. Figure ~\ref{fig:funcionamento_quebra_linearidade} illustrates the algorithm.
